Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-13 Manuscript/manuscript.Rmd
combining arrowhead and peakachu; methods edits

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *P*, to consider at each node. The number of trees that were aggregated for each RF model was set to the default, $n=500$. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and loop boundary calling algorithms, we applied our RF model to predict boundaries from base pair (bp) resolution data. [??? Write how your vector Y and annotation matrices change;!!!added ???You describe whole-genome model. It should have been described early on. Here, describe chromosome-specific bp-resolution model]. At this point, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements). Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. Given that each chromosome is composed of tens, in some cases hundreds, of millions of base pairs, we used the same whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

Naturally, when dealing with such fine resolution, there will be spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances). To guard against inflating the number of predicted TAD boundaries we developed a technique that combines both density-based spatial clustering (DBSCAN) and partitioning techniques to identify precise TAD boundaries throughout the linear genome.

First, we extracted the vector $p_{n_{i}}$ that denoted the probability of being a boundary point for each chromosome-specific base pair from our RF model, where $n_{i}$ was the length of chromosome $i$. We then applied *DBSCAN* (version 1.1-5) to the matrix of pairwise distances [??? Is that true? Matrix is 2-dimensional] between base pairs with $p_{n_{i}}=1$. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts* [@hahsler2019dbscan]. The eps parameter was assigned as the respective data resolution [??? Meaning?]. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3 [??? Meaning?]. The resulting clusters of density-reachable base pairs identified by DBSCAN were termed *predicted TAD boundary regions* (PTBR) [??? Use simple words: "boundary region" (many bases clustered by DBSCAN), then "boundary base" (selected by PAM)]. To precisely identify the most representative base pair among each PTBR we performed an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base pair within each clustered PTBR.

@@

where A and B represent genomic regions created by flanking called and predicted boundaries.

@@

We hypothesized that the type of genomic annotations may affect the predictive performance. Using the aforementioned optimal settings, we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, and these results were consistent for loop boundaries (Figure 4A; Supplementary Figures 4A,5A,6A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines.

@@

We sought to further optimize our TAD prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B,5B,6B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary predigtion, we clustered the predictive importances of top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C,5C,6C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self].

@@

![**Assessing top ranked transcription factors in TAD boundary prediction for GM12878.** (A) Barplots comparing performances of models using specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (BroadHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model containing all three classes (ALL; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/figure4_annots_specifc_rfe_importances.png)

@@

Here, we opted to implement preciseTAD using an optimally built whole genome random forest model. That is, a random forest model built on 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143). The reason for this was because of the improved performance that was exhibited for a whole genome random forest compared to aggregated chromosome-specific models (Supplementary Figure 7).

@@

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD boundary prediction for K562.** (A) Barplots comparing performances of models built using ARROWHEAD boundaries on specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (BroadHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model containing all three classes (ALL; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.

**Supplementary Figure 5. Assessing top ranked transcription factors in loop boundary prediction for GM12878.** (A) Barplots comparing performances of models built using PEAKACHU boundaries on specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (BroadHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model containing all three classes (ALL; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.

**Supplementary Figure 6. Assessing top ranked transcription factors in loop boundary prediction for K562.** (A) Barplots comparing performances of models built using PEAKACHU boundaries on specific classes of functional genomic elements as the feature space including histone modifications (HM; red), cis-regulated elements (BroadHMM; green), transcription factor binding sites (TFBS; blue), in addition to a model containing all three classes (ALL; purple). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.

>>>>

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, predictor types, and re-sampling techniques. Five-fold cross validation was used to tune the number of annotations, *P*, to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to 1. The performance of each model was compared using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

To investigate whether we could alleviate resolution limitations of conventional TAD and chromatin  loop calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base pair (bp) resolution. To demonstrate, we opted to implement an optimally built whole genome RF model. That is, a random forest model built on the optimal combination of data resolution (binning), predictor type (feature space), re-sampling technique (class balancing correction), and annotations (most informative genomic elements) [??? Write how your vector Y and annotation matrices change;!!!added ???You describe whole-genome model. It should have been described early on. Here, describe chromosome-specific bp-resolution model]. Here, $Y$ was of length $N$ where $N=n_1+n_2+\cdots+n_{22}$, the total number of bins across each chromosome. The data matrix $\textbf{X}$ had dimensions $N \times p_s$, where $p_s \subset P$, a subset of the most predictive genomic elements. The whole genome RF was tuned sequentially. First, *mtry*=$\sqrt{p}$ and *nodesize*=1, then we allowed *ntree* to range from 10 to 1000 by values of 10, and determined where the balanced accuracies converged, say at $t$. Next, we set *mtry*=$\sqrt{p}$ and *ntree*=$t$ and allowed *nodesize* to range from 1 to 21 by values of 2, again evaluating where performances converged. <!--Given that each chromosome is composed of tens, in some cases hundreds, of millions of base pairs,--> We then used the same whole genome RF model to predict boundaries from chromosome-specific bp resolution data.

Naturally, when dealing with such fine resolution, there will be spatially close base pairs with similar predictor profiles (i.e. similar overlap counts, distances, etc.). *preciseTAD* guards against inflating the number of predicted boundaries by combining both density-based spatial clustering (DBSCAN) and partitioning techniques to identify precise TAD (or loop) boundaries throughout the linear genome. First the vector, $p_{n_{i}}$, denoting each base pair's probability of being a boundary point from the RF model, was extracted, where $n_{i}$ represents the length of chromosome $i$. Next, *DBSCAN* (version 1.1-5) was applied to the matrix of pairwise genomic distances [??? Is that true? Matrix is 2-dimensional] between base pairs with $p_{n_{i}}=1$. The DBSCAN algorithm takes two parameters, *eps* ($\epsilon$) and *MinPts* [@hahsler2019dbscan]. The eps parameter was assigned as the respective data resolution [??? Meaning?]. The heuristic being that base pairs with genomic distances less than resolution should occupy the same designated cluster. The corresponding MinPts parameter was set to 3 [??? Meaning?]. The resulting clusters of density-reachable base pairs identified by DBSCAN were termed *preciseTAD boundary regions* (PTBR) [??? Use simple words: "boundary region" (many bases clustered by DBSCAN), then "boundary base" (selected by PAM)]. To precisely identify a single base pair among each PTBR, *preciseTAD* then implements an extension of partitioning around medoids (PAM), known as CLARA (Clustering Large Applications) for better scaling and more computational efficiency [@reynolds2006clustering; @wei2000empirical]. Thusly, the corresponding cluster medoid was defined as a *preciseTAD boundary point* (PTBP), making it the most representative base pair within each clustered PTBR.

@@

where A and B represent genomic regions created by flanking called and predicted boundaries. That is, the set of shared boundaries divided by the total number of unique boundaries.

@@

We hypothesized that the type of genomic annotations may affect the predictive performance. Using the aforementioned optimal settings, we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, and these results were consistent for loop boundaries, for both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both the GM12878 and K562 cell lines.

@@

We sought to further optimize our TAD prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self].

@@

![**Assessing top ranked transcription factors in TAD and loop boundary prediction for GM12878.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using specific classes of functional genomic elements as the feature space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.](Figures/figure4_annots_specifc_rfe_importances_arrowhead_peakachu.png)

@@

Here, we opted to implement preciseTAD using an optimally built whole genome random forest model. That is, a random forest model built on 5 kb defined genomic bins, with random under-sampling, and distance-type predictors for the top informative TFBS (CTCF, RAD21, SMC3, and ZNF143). The reason for this was because of the improved performance that was exhibited for a whole genome random forest compared to aggregated chromosome-specific models (Supplementary Figure 7). When tuning the full genome RF we found that performances converged when *mtry*=2, *ntree*=100, and *nodesize*=???.

@@

**Supplementary Figure 4. Assessing top ranked transcription factors in TAD and loop boundary prediction for K562.** (A) Barplots comparing performances of TAD (ARROWHEAD) and loop (PEAKACHU) boundary prediction models using specific classes of functional genomic elements as the feature space including histone modifications (HM), cis-regulated elements (BroadHMM), transcription factor binding sites (TFBS), in addition to a model containing all three classes (ALL). (B) Recursive feature elimination (RFE) analysis used to reduce the number of TFBS to include in the optimally built random forest model. Error bars represent standard deviation from the cross-validated accuracy across each chromosome. (C) Clustered heatmap of predictive importances for the union of the top 8 most predictive TFBS across chromosomes as determined from performing RFE.
